---
title: "Statistical Analysis Report"
author: "Molly McNamara"
output:
  html_document:
    fig_caption: yes
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
## Load data packages that are needed
library(dplyr)
library(ggplot2)
library(readr)
library(countrycode)
library(knitr)

## Import the data.  Fix an extra column that keeps importing from the CSV. Add a birth region converted from birth country.
Hockey <- read_csv("C:/Users/mmcnamara/Desktop/FinalData_clean.csv")
Hockey <- select(Hockey, -X1)
Hockey$BirthRegion <- countrycode(Hockey$Birth_Country, 'country.name', 'continent', warn = TRUE, custom_dict = NULL,
                                  custom_match = NULL, origin_regex = FALSE)
## Create a subset that only includes the first instance of each player. This way analysis of certain attributes is not skewed by multiple seasons of play by a player.
ByPlayer <- Hockey[match(unique(Hockey$Player_ID), Hockey$Player_ID),]
```

## Initial Evaluation of the Dataset

The datset was initially evaluated with the summary command to review the variables and their basic statistics.

```{r , eval=TRUE}
## Look at the main dataset.
summary(Hockey)
```

Basic statistics about players and their draft information were viewed next.
```{r, eval = TRUE}
summary(ByPlayer[,c("Height", "Weight", "FirstNHLseason", "LastNHLseason", "Draft_Pick","Draft_Year", "Draft_Age", "Draft_Round", "Draft_Pick")])

```

## Draft Statistics

Draft statistics were initially evaluated by viewing a distribution of draft position of the players in this dataset.
```{r, echo = FALSE}
ggplot(ByPlayer, aes(x = Draft_Round)) + geom_bar()
ggplot(ByPlayer, aes(x = Draft_Pick)) + geom_histogram()
```

Of the players in these years 2000-2012, it appears to be skewed in one direction, towards players drafted in earlier rounds.  It may be that players drafted in earlier rounds play longer in the league and therefore are more highly represented.

To further investigate that point, we make a comparison between draft round and years in the NHL for players in this dataset.
```{r, echo = FALSE}
YearsPlayed = ByPlayer$LastNHLseason - ByPlayer$FirstNHLseason
ggplot(ByPlayer, aes(x = Draft_Round, y = YearsPlayed)) + geom_jitter() + geom_smooth(method = lm)
```
It does indeed appear that players from earlier draft rounds may have somewhat longer NHL careers.

To further explore factors around the NHL draft, tables were generated that summarized what round a player was drafted in and at what age by birth country.

```{r kable, echo=FALSE}
Draftroundbycountry <- ByPlayer %>% group_by(Birth_Country) %>% summarize(mean=mean(Draft_Round))
Draftagebycountry <- ByPlayer %>% group_by(Birth_Country) %>% summarize(mean=mean(Draft_Age))
knitr::kable(Draftroundbycountry)
knitr::kable(Draftagebycountry)
```
Draft age looks fairly consistent, though this can be examined further later.

Next is a simple comparison of how many players were drafted from different geographical regions.
```{r, echo=FALSE}
ggplot(ByPlayer, aes(x = BirthRegion)) + geom_histogram(stat="count")
```
Clearly the Americas source most NHL players followed by Europe.

One might wonder if players from some regions drafted earlier or later in the draft on average.
```{r, echo=FALSE}
ggplot(ByPlayer, aes(BirthRegion, Draft_Round)) + geom_boxplot()
```
There may be some differences between regions, to be examined further later. As seen in the prior graph, the majority of players come from the Americas and Europe, which appear to be distributed similarly here.

## Scoring Statistics

Shots per game and penalty minutes are player statistics that might impact their goal scoring.
```{r, echo=FALSE}
ggplot(Hockey, aes(x = ShotsPerGame, y = Goals)) + geom_point()
ggplot(Hockey, aes(x = Penalty_Minutes, y = Goals)) + geom_point()
```

As one might expect, it appears the more shots a player takes, the more goals he scores.The relationship between time spent in the penalty box and goals is less clear (one might expect more time in the box leaves less time for goals).

It might be expected that teams who score more goals in a year are more likely to go to the playoffs.
```{r, echo = FALSE}
ggplot(Hockey, aes(x = Year, y=Team_GoalsFor, col = Playoffs)) + geom_point()
```
This graph seems to indicate such, though there are cases where the opposite is true.

Do players who are drafted in earlier rounds score a greater percentage of their teams' goals?
```{r}
ggplot(Hockey, aes(x = Year, y=PercentGoals, col = Draft_Round)) + geom_jitter()
```
It is difficult to tell from this graph. but maybe?

Do players drafted in earlier rounds score more goals per game? Does this affect their team making the playoffs? Is there any difference based on their region of origin?
```{r, echo = FALSE}
ggplot(Hockey, aes(x = Draft_Pick, y=GoalsPerGame, col=Playoffs)) + geom_point(alpha = 0.5, position = "jitter") + geom_smooth(method = lm)
ggplot(Hockey, aes(x = Draft_Pick, y=GoalsPerGame, col=BirthRegion)) + geom_point(alpha = 0.5) 
```
It does look like perhaps the trend is for more goals per game from higher-drafted players but it's not clear if this impacts making the playoffs.  The impact of birth region is not entirely clear though it seems possibly higher-drafted European players score a greater number of goals per game.

Does draft age impact goal scoring?
```{r, echo = FALSE}
posn.jd <- position_jitterdodge(0.5, 0, 0.6)
ggplot(Hockey, aes(x = Draft_Age, y = Goals, color = Draft_Round)) + 
  geom_point(size = 3, alpha = 0.5, position = posn.jd) + 
  facet_grid(. ~ Playoffs)
```

Players drafted younger may score more goals, on teams that do and don't make the playoffs, if they were players drafted in early rounds.

Does there appear to be any difference in goal scoring, as a function of round drafted, when compared between teams that do and don't make the playoffs and looking at birth region?
```{r, echo = FALSE}
ggplot(Hockey, aes(x = Draft_Round, y = Goals, color = BirthRegion)) + 
  geom_point(size = 3, alpha = 0.5, position = posn.jd) + 
  facet_grid(. ~ Playoffs)
```
No clear difference is seen.


